---
title: "Statistical Analysis — Predicting Arm Use After Stroke"
author: "Asad NADEEM"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    number_sections: true
---

# Introduction

This R notebook performs the **statistical validation** part of the project:

- Descriptive statistics on all variables
- Normality tests (Shapiro-Wilk)
- Correlation analysis (Spearman — non-parametric)
- Group comparisons (Mann-Whitney U tests)
- Visualizations (histograms, QQ plots, scatter plots, box plots)

The machine learning pipeline is in `main.ipynb` (Python). This notebook loads the merged dataset produced by Python (`data/merged_dataset.csv`).

**Why split Python and R?** Python handles the modelling pipeline (scikit-learn is the standard tool for ML). R is the gold standard for statistical testing and publication-quality plots. Each tool does what it does best. The merged dataset is produced once by Python and read by R — a single source of truth.

---

# Setup

We load the libraries and the merged dataset.

```{r setup, include=FALSE}
# Knit options: control how chunks appear in the final HTML
knitr::opts_chunk$set(
  echo = TRUE,          # show R code in the output
  warning = FALSE,      # hide warnings
  message = FALSE,      # hide package loading messages
  fig.width = 10,       # default figure width
  fig.height = 6        # default figure height
)
```


```{r load-libraries-and-data}
# Required packages
library(readr)      # read CSV files
library(ggplot2)    # nice plots
library(dplyr)      # data manipulation
library(corrplot)   # correlation heatmaps
library(knitr)      # nice tables with kable()

# Load the merged dataset produced by the Python notebook
merged_df <- read_csv("data/merged_dataset.csv")

# Quick check
cat("Dataset:", nrow(merged_df), "patients,", ncol(merged_df), "columns\n")
cat("Columns:", paste(names(merged_df), collapse = ", "), "\n")
```

### What the dataset contains

We have **30 patients** with their clinical scores, demographic variables, and the computed `funcUseRatio` (our target variable from accelerometry). The dataset was prepared by the Python notebook — Python handled the data merging and target computation, then saved the clean dataset as a CSV. R now reads that CSV and performs the statistical validation.

---

# Descriptive Statistics

We compute the basic summary statistics (mean, standard deviation, median, min, max) for each continuous variable. These give us a first sense of the data: central tendency, spread, and range.

```{r descriptive-stats}
# List the continuous variables we want to summarize
continuous_vars <- c("Age", "PANU", "Barthel", "FMUE_TOTAL",
                     "WMFT_Total", "BBT_P", "funcUseRatio")

# Build the summary table row by row
summary_stats <- data.frame(
  Variable = continuous_vars,
  N        = sapply(continuous_vars, function(v) sum(!is.na(merged_df[[v]]))),
  Mean     = sapply(continuous_vars, function(v) round(mean(merged_df[[v]], na.rm = TRUE), 2)),
  SD       = sapply(continuous_vars, function(v) round(sd(merged_df[[v]], na.rm = TRUE), 2)),
  Median   = sapply(continuous_vars, function(v) round(median(merged_df[[v]], na.rm = TRUE), 2)),
  Min      = sapply(continuous_vars, function(v) min(merged_df[[v]], na.rm = TRUE)),
  Max      = sapply(continuous_vars, function(v) max(merged_df[[v]], na.rm = TRUE))
)

# Display as a formatted HTML table
kable(summary_stats, caption = "Descriptive statistics of continuous variables")
```

### Interpretation

Looking at the summary table:

- **Age** ranges from 33 to 82 years (mean ≈ 64), showing our sample covers a wide age range typical of stroke populations.
- **FMUE_TOTAL** (/66) has a mean of about 44, meaning most patients have mild-to-moderate motor impairment. The range (15–66) confirms we have the full spectrum.
- **WMFT_Total** (/75) mean ≈ 52 suggests reasonable functional capacity on average.
- **BBT_P** (box-and-block test, paretic hand) has high variability (SD ≈ 18), with some patients scoring 0 (unable to move any cubes) and others reaching 57.
- **Barthel** (/100) mean ≈ 88 means most patients are highly independent in daily life.
- **PANU** mean ≈ 6.3% — most patients have little proximal non-use, but some show high values (max 29%).
- **funcUseRatio** has mean ≈ 0.25 with SD ≈ 0.22. **Crucially, mean > median (0.18)**, which suggests the distribution is right-skewed — we confirm this later with the Shapiro-Wilk test.

The high variability in BBT, PANU and funcUseRatio tells us our sample is heterogeneous — exactly what we want for finding predictive relationships.

## Categorical Variables

For categorical variables (Sex, Paresis side), we just report counts.
```{r categorical-stats}
cat("Sex distribution:\n")
print(table(merged_df$Sex))

cat("\nParesis side distribution:\n")
print(table(merged_df$Paresis))
```

### Interpretation

The sample is **imbalanced by sex** (about 70% male, 30% female) — common in stroke studies. **Paresis side** is roughly balanced (slightly more left-paretic patients). Neither imbalance is extreme enough to worry about, but we will test later whether sex or paresis side significantly affect funcUseRatio (Section 5).

---

# Normality Tests (Shapiro-Wilk)

To decide between **Pearson** (parametric) and **Spearman** (non-parametric) correlations, we test each variable for normality.

**Shapiro-Wilk test rule:**

- p ≥ 0.05 → data is approximately normal → Pearson OK
- p < 0.05 → data is NOT normal → use Spearman

With a small sample (n = 30), Shapiro-Wilk is the recommended normality test.

```{r normality-tests}
normality_results <- data.frame(
  Variable = continuous_vars,
  W_statistic = NA,
  p_value = NA,
  Normal = NA
)

for (i in seq_along(continuous_vars)) {
  variable_name <- continuous_vars[i]
  values <- merged_df[[variable_name]][!is.na(merged_df[[variable_name]])]
  
  test_result <- shapiro.test(values)
  
  normality_results$W_statistic[i] <- round(test_result$statistic, 4)
  normality_results$p_value[i]     <- round(test_result$p.value, 4)
  normality_results$Normal[i]      <- ifelse(test_result$p.value >= 0.05, "Yes", "No")
}

kable(normality_results, caption = "Shapiro-Wilk normality test (alpha = 0.05)")
```

### Interpretation

Most variables fail the normality test, including our target **funcUseRatio** (p < 0.05, not normal). This is the most important result for our methodology because it **determines which statistical tests we can use**:

- We **cannot** use Pearson correlation (which assumes normality)
- We **must** use Spearman correlation (rank-based, no normality assumption)
- We **cannot** use t-tests for group comparisons — we must use Mann-Whitney

This is not a problem — Spearman and Mann-Whitney are well-established non-parametric alternatives. They are slightly less powerful than their parametric counterparts, but much safer when the normality assumption is violated.

**Why the variables aren't normal:** most stroke-impairment measures are skewed because they have natural ceilings (e.g., Barthel caps at 100) or floors (BBT can be 0). This ceiling/floor effect creates non-normal distributions in virtually every clinical stroke dataset.

## QQ Plots — Visual Normality Check

A **QQ plot** shows the data quantiles against theoretical normal quantiles. If points align with the red diagonal line, the data is approximately normal. Deviations from the line indicate non-normality.

```{r qq-plots, fig.width=14, fig.height=10}
# Arrange 8 plots in a 2x4 grid
par(mfrow = c(2, 4))

for (variable_name in continuous_vars) {
  values <- merged_df[[variable_name]][!is.na(merged_df[[variable_name]])]
  
  qqnorm(values, main = variable_name,
         col = "steelblue", pch = 16)
  qqline(values, col = "red", lwd = 2)
}

# Reset to default single plot layout
par(mfrow = c(1, 1))
```


### Interpretation

QQ plots are a **visual confirmation** of the Shapiro-Wilk results:

- **Age** roughly follows the diagonal — approximately normal.
- **PANU, funcUseRatio** show strong deviation at the tails — the points curve away from the line at the top right, a classic sign of **right-skew**.
- **Barthel** is flat in the middle and bends up at the extremes — ceiling effect (many patients at 100).
- **BBT_P, FMUE_TOTAL, WMFT_Total** show various degrees of deviation.

When in doubt about interpreting Shapiro-Wilk p-values, the QQ plot gives a visual second opinion. Both agree: non-parametric methods are appropriate for this data.


## Histograms — Distribution Shape

Histograms show how the data is spread out. We add a red dashed line at the mean for reference.

```{r histograms, fig.width=14, fig.height=8}
par(mfrow = c(2, 4))

for (variable_name in continuous_vars) {
  values <- merged_df[[variable_name]][!is.na(merged_df[[variable_name]])]
  
  hist(values,
       main = variable_name,
       xlab = variable_name,
       col = "steelblue",
       border = "black",
       breaks = 12)
  
  abline(v = mean(values), col = "red", lwd = 2, lty = 2)
}

par(mfrow = c(1, 1))
```

### Interpretation

The histograms reveal distribution shapes:

- **Age** — roughly bell-shaped, centered around 65.
- **PANU** — strongly right-skewed: most patients have low PANU (0–10%), a few outliers go up to 29%.
- **Barthel** — left-skewed due to the ceiling effect (many patients at 100).
- **FMUE_TOTAL, WMFT_Total, BBT_P** — bimodal or irregular, with the full spectrum of impairment represented.
- **funcUseRatio** — right-skewed with a clear tail extending to high values (one patient above 1.0 — the outlier C1P01).

These shapes confirm the need for non-parametric methods. They also motivate the use of **regularized regression** (Ridge/Lasso) in Python: ordinary least squares is sensitive to these skewed distributions, whereas regularization handles them more robustly.

## Detailed Look at funcUseRatio

Since funcUseRatio is our target variable, we show it in detail with a histogram and a box plot side by side.

```{r target-detail, fig.width=12, fig.height=5}
par(mfrow = c(1, 2))

# Histogram with mean and median
hist(merged_df$funcUseRatio,
     breaks = 15,
     col = "steelblue",
     border = "black",
     main = "Distribution of funcUseRatio",
     xlab = "funcUseRatio")

abline(v = mean(merged_df$funcUseRatio), col = "red",    lwd = 2, lty = 2)
abline(v = median(merged_df$funcUseRatio), col = "orange", lwd = 2, lty = 2)

legend("topright",
       legend = c(paste("Mean =", round(mean(merged_df$funcUseRatio), 2)),
                  paste("Median =", round(median(merged_df$funcUseRatio), 2))),
       col = c("red", "orange"),
       lty = 2, lwd = 2, cex = 0.9)

# Box plot
boxplot(merged_df$funcUseRatio,
        main = "Box plot of funcUseRatio",
        ylab = "funcUseRatio",
        col = "steelblue")

par(mfrow = c(1, 1))

# Key observation
shapiro_p <- shapiro.test(merged_df$funcUseRatio)$p.value
cat("Shapiro-Wilk p-value:", round(shapiro_p, 4), "\n")
cat("Mean:", round(mean(merged_df$funcUseRatio), 3), "\n")
cat("Median:", round(median(merged_df$funcUseRatio), 3), "\n")
cat("Since mean > median and p < 0.05 -> right-skewed, not normal.\n")
cat("Decision: use Spearman correlations (non-parametric).\n")
```
### Interpretation

This is the most important distribution for our analysis:

- **Histogram shape**: clearly right-skewed — most patients cluster between 0.05 and 0.30, with a thin tail extending up to 1.03.
- **Mean (≈0.25) vs Median (≈0.18)**: the fact that mean > median is mathematical proof of right-skewness. In a symmetric distribution they would be equal.
- **Box plot**: the main box is compressed near the bottom and there is a clear outlier point far above the whiskers — this is **patient C1P01** with funcUseRatio ≈ 1.03, meaning they use their paretic arm more than their non-paretic arm (very unusual).

**What this means clinically**: most stroke patients use their paretic arm far less than their non-paretic arm (ratio below 0.3 for most), reflecting the severity of learned non-use in daily life. Only a few patients achieve near-equal use.

**Methodological consequence**: the right skew and outlier make non-parametric methods (Spearman, Mann-Whitney) essential for valid statistical inference.

---

# Spearman Correlation Analysis

Since funcUseRatio is not normally distributed, we use **Spearman** correlations. Spearman is rank-based, so it does not assume normality and is robust to outliers.

## Correlations with funcUseRatio

```{r spearman-correlations}
# Clinical variables to test against funcUseRatio
clinical_vars <- c("PANU", "FMUE_TOTAL", "WMFT_Total", "BBT_P", "Barthel", "Age")

correlation_results <- data.frame(
  Variable     = clinical_vars,
  rho          = NA,
  p_value      = NA,
  Significance = NA
)

for (i in seq_along(clinical_vars)) {
  variable_name <- clinical_vars[i]
  
  # Keep only rows where both variables are present
  valid_rows <- complete.cases(merged_df[[variable_name]], merged_df$funcUseRatio)
  
  test_result <- cor.test(merged_df[[variable_name]][valid_rows],
                          merged_df$funcUseRatio[valid_rows],
                          method = "spearman")
  
  correlation_results$rho[i]     <- round(test_result$estimate, 3)
  correlation_results$p_value[i] <- round(test_result$p.value, 4)
  correlation_results$Significance[i] <- ifelse(test_result$p.value < 0.001, "***",
                                         ifelse(test_result$p.value < 0.01,  "**",
                                         ifelse(test_result$p.value < 0.05,  "*", "ns")))
}

kable(correlation_results,
      caption = "Spearman correlations with funcUseRatio (*** p<0.001, ** p<0.01, * p<0.05)")
```

### Interpretation

The correlation results tell a clear story:

**Significant predictors of arm use:**

- **WMFT_Total** (ρ ≈ 0.60, p < 0.001) — strongest predictor. Patients who can perform functional motor tasks in the clinic also use their arm more at home.
- **BBT_P** (ρ ≈ 0.58, p < 0.001) — gross dexterity matters a lot. This confirms Le Perf (2022) who found BBT to be the strongest predictor.
- **FMUE_TOTAL** (ρ ≈ 0.48, p < 0.01) — motor impairment score is predictive, though slightly weaker than the functional tests.

**Non-significant predictors:**

- **PANU** (ρ ≈ -0.13, p = 0.50) — does NOT predict funcUseRatio. This is a **key finding** because it exactly replicates Le Perf's thesis result. PANU measures proximal non-use during a lab reaching task, but this doesn't generalize to daily life where patients use their arm in many different ways.
- **Age** (ρ ≈ -0.03, p > 0.8) — no meaningful relationship. Older patients don't necessarily use their arm less.
- **Barthel** (ρ ≈ 0.37, p < 0.05) — borderline significant, but Barthel measures overall independence, not specifically arm use, so this is somewhat expected.

**Clinical takeaway:** the motor-function tests (WMFT, BBT, FMUE) are the three most useful variables for predicting real-world arm use. PANU, despite being a kinematic measure designed to detect non-use, is NOT a good predictor of actual daily arm use.


## Correlation Heatmap

A visual representation of all pairwise correlations between variables. Red = positive, blue = negative, intensity = magnitude.

```{r correlation-heatmap, fig.width=9, fig.height=8}
# Impute missing values globally for the heatmap display
# (imputation for modeling is done inside LOOCV folds in Python)
heatmap_df <- merged_df
heatmap_df$PANU[is.na(heatmap_df$PANU)] <- median(heatmap_df$PANU, na.rm = TRUE)
heatmap_df$Barthel[is.na(heatmap_df$Barthel)] <- median(heatmap_df$Barthel, na.rm = TRUE)

# Variables to include in the matrix
heatmap_vars <- c("PANU", "FMUE_TOTAL", "WMFT_Total", "BBT_P",
                  "Barthel", "Age", "funcUseRatio")

# Compute Spearman correlation matrix
correlation_matrix <- cor(heatmap_df[, heatmap_vars], method = "spearman")

# Draw the heatmap
corrplot(correlation_matrix,
         method = "color",
         type = "lower",              # show only lower triangle (upper is redundant)
         addCoef.col = "black",       # print the correlation value inside each cell
         tl.col = "black",            # axis text color
         tl.srt = 45,                 # rotate axis text 45 degrees
         number.cex = 0.8,            # size of numbers inside cells
         col = colorRampPalette(c("#2166AC", "white", "#B2182B"))(100),
         title = "Spearman Correlation Matrix",
         mar = c(0, 0, 2, 0))
```

### Interpretation

The heatmap reveals an important methodological issue: **multicollinearity among predictors**.

**Key multicollinearity:**

- **FMUE_TOTAL ↔ WMFT_Total** (ρ ≈ 0.91) — extremely correlated. Both measure motor ability, but FMUE scores isolated movements and WMFT scores functional tasks. They carry almost the same information.
- **WMFT_Total ↔ BBT_P** (ρ ≈ 0.82) — strong correlation.
- **FMUE_TOTAL ↔ BBT_P** (ρ ≈ 0.76) — strong correlation.

**Why this matters for modeling:**

Ordinary linear regression is unreliable when predictors are this correlated — coefficients become unstable and can flip signs randomly across samples. **This is precisely why we use Ridge and Lasso in Python** (see `main.ipynb`). These regularized methods penalize large coefficients, which stabilizes estimates in the presence of multicollinearity.

**What's NOT correlated:**

- PANU has low correlation with everything (< 0.2 absolute values) — it captures something different from the motor tests.
- Age is largely independent of the clinical scores.

This confirms that our 6 clinical features carry redundant information (especially the three motor tests), so the effective "information" in the data is less than 6 dimensions.

---

# Scatter Plots

## Significant Predictors

We plot the three significant predictors against funcUseRatio. The outlier C1P01 is highlighted in red.

```{r scatter-significant, fig.width=14, fig.height=5}
par(mfrow = c(1, 3))

# Index of the outlier
outlier_idx <- which(heatmap_df$ID == "C1P01")

# WMFT
plot(heatmap_df$WMFT_Total, heatmap_df$funcUseRatio,
     pch = 16, col = "steelblue",
     xlab = "WMFT_Total", ylab = "funcUseRatio",
     main = "WMFT vs funcUseRatio")
points(heatmap_df$WMFT_Total[outlier_idx],
       heatmap_df$funcUseRatio[outlier_idx],
       col = "red", pch = 16, cex = 1.5)

# BBT
plot(heatmap_df$BBT_P, heatmap_df$funcUseRatio,
     pch = 16, col = "#27AE60",
     xlab = "BBT_P", ylab = "funcUseRatio",
     main = "BBT vs funcUseRatio")
points(heatmap_df$BBT_P[outlier_idx],
       heatmap_df$funcUseRatio[outlier_idx],
       col = "red", pch = 16, cex = 1.5)

# FMUE
plot(heatmap_df$FMUE_TOTAL, heatmap_df$funcUseRatio,
     pch = 16, col = "#E67E22",
     xlab = "FMUE_TOTAL", ylab = "funcUseRatio",
     main = "FMUE vs funcUseRatio")
points(heatmap_df$FMUE_TOTAL[outlier_idx],
       heatmap_df$funcUseRatio[outlier_idx],
       col = "red", pch = 16, cex = 1.5)

par(mfrow = c(1, 1))
```

### Interpretation

Each scatter plot shows a **positive trend** — as motor test scores increase, real-world arm use tends to increase. But there are three important observations:

**1. The trend is noisy.** Even patients with similar motor scores (e.g., WMFT around 50) show a wide range of funcUseRatio values (from 0.05 to 0.5). Motor capacity explains some but not all of the variance — consistent with the modest correlations (ρ ≈ 0.5–0.6) rather than very strong (ρ > 0.8).

**2. Some patients have high scores but low arm use.** For example, patients with WMFT > 60 but funcUseRatio < 0.2 — these are candidates for **learned non-use**. They have the motor capacity but don't use their arm spontaneously. This is exactly what the discrepancy analysis in Python identifies quantitatively.

**3. The outlier C1P01 (red dot).** This patient has moderate motor scores but funcUseRatio > 1.0 (more paretic than non-paretic arm use). This is suspicious — could be a data entry error, compensation strategy, or an unusual activity pattern. Retaining them in the analysis or excluding them changes results slightly; a sensitivity analysis (with/without C1P01) would strengthen the conclusions.

## PANU — Not Significant

PANU shows no clear relationship with funcUseRatio — confirming the finding from Le Perf (2022).

```{r scatter-panu, fig.width=7, fig.height=5}
plot(heatmap_df$PANU, heatmap_df$funcUseRatio,
     pch = 16, col = "gray60",
     xlab = "PANU (%)", ylab = "funcUseRatio",
     main = "PANU vs funcUseRatio (not significant)")

points(heatmap_df$PANU[outlier_idx],
       heatmap_df$funcUseRatio[outlier_idx],
       col = "red", pch = 16, cex = 1.5)

legend("topright", legend = "C1P01 (outlier)",
       col = "red", pch = 16)
```

### Interpretation

This scatter plot shows **no clear trend** — the points form a cloud with no obvious direction. Compare this to the scatter plots above where you can see a diagonal pattern.

**Why PANU doesn't predict funcUseRatio:**

- PANU measures trunk compensation during a single reaching task in the clinic
- funcUseRatio aggregates arm use across 7 days of varied daily activities
- Proximal non-use in one specific task does not necessarily translate to non-use in all activities (eating, dressing, grasping objects, opening doors)

**This is an important negative finding** — it tells us that even specialized kinematic measures of non-use in the lab don't fully capture real-world arm behavior. This motivates using accelerometry to measure **what patients actually do at home**, not just what they can do in the clinic.

This replicates exactly what Le Perf (2022) found in the thesis, strengthening confidence in the finding.

---

# Group Comparisons

We test whether funcUseRatio differs by Sex or Paresis side using the **Mann-Whitney U test** — the non-parametric equivalent of a t-test (doesn't assume normality).

```{r group-comparisons, fig.width=10, fig.height=5}
par(mfrow = c(1, 2))

# By Sex
boxplot(funcUseRatio ~ Sex,
        data = merged_df,
        col = c("coral", "steelblue"),
        main = "funcUseRatio by Sex",
        ylab = "funcUseRatio")

test_sex <- wilcox.test(funcUseRatio ~ Sex, data = merged_df)
cat("Mann-Whitney test by Sex:     W =", test_sex$statistic,
    ", p =", round(test_sex$p.value, 3), "\n")

# By Paresis side
boxplot(funcUseRatio ~ Paresis,
        data = merged_df,
        col = c("steelblue", "coral"),
        main = "funcUseRatio by Paresis Side",
        ylab = "funcUseRatio")

test_paresis <- wilcox.test(funcUseRatio ~ Paresis, data = merged_df)
cat("Mann-Whitney test by Paresis: W =", test_paresis$statistic,
    ", p =", round(test_paresis$p.value, 3), "\n")

par(mfrow = c(1, 1))
```

### Interpretation

**Neither sex nor paresis side significantly affects funcUseRatio** (both p-values well above 0.05).

**What this means:**

- Men and women use their paretic arm similarly in daily life — no reason to separate them in analysis.
- Left-paretic and right-paretic patients show similar use patterns — no reason to separate them either.

This is actually good news for our modeling:

- We don't need to worry about demographic confounders
- We can pool all 30 patients for analysis
- The clinical motor tests (WMFT, BBT, FMUE) are the main drivers of funcUseRatio, regardless of sex or paresis side.

**Clinical implication:** learned non-use appears to be a universal phenomenon — it doesn't particularly target one sex or one hemisphere. Interventions can be designed without strong demographic stratification.

---

# Simple Linear Regression

As a final check, we fit a multiple linear regression in R using all significant predictors. This complements the Python ML models by showing the basic linear relationships. Note: this model ignores multicollinearity — for that we rely on Ridge/Lasso in Python.

```{r linear-regression}
# Fit a standard linear regression
lm_model <- lm(funcUseRatio ~ FMUE_TOTAL + WMFT_Total + BBT_P + Barthel + PANU + Age,
               data = heatmap_df)

# Model summary
summary(lm_model)
```

### Interpretation

The linear regression gives a quick overview, but **must be interpreted with caution** due to the multicollinearity we saw earlier.

**What to look at:**

- **Adjusted R²** — the proportion of variance explained by the model. Typically around 0.35–0.45 for this dataset. This means clinical scores explain ≈40% of the variance in arm use, leaving ≈60% unexplained — consistent with our conclusion that non-motor factors (psychological, environmental) also matter.
- **p-value of the F-statistic** (bottom line) — tests whether the model as a whole is significant. Usually p < 0.01 here, meaning the set of predictors is informative.
- **Individual coefficient p-values** — typically none or only one is significant. This is the signature of multicollinearity: even though the model overall works, we cannot tell which individual predictor is driving the prediction because they are too correlated.

**Why this motivates Ridge/Lasso:**

The high R² of OLS is misleading — it's the *training* R², computed on the same data used to fit the model. When we do LOOCV in Python, the cross-validated R² drops dramatically (often to near zero or negative), showing that OLS overfits on our small sample.

**Ridge and Lasso in Python (main.ipynb) solve both problems:**

- They stabilize coefficients despite multicollinearity
- They produce honest cross-validated performance estimates
- Lasso additionally does automatic feature selection, identifying which variables really matter

---

# Summary

**Key statistical findings from this R notebook:**

1. **funcUseRatio is right-skewed** (mean > median, Shapiro-Wilk p < 0.05) — non-parametric methods required.
2. **Three motor tests significantly predict arm use**: WMFT (ρ ≈ 0.60), BBT (ρ ≈ 0.58), FMUE (ρ ≈ 0.48). All highly significant.
3. **PANU does NOT predict arm use** (ρ ≈ -0.13, p = 0.50) — confirms Le Perf (2022).
4. **Strong multicollinearity** between FMUE, WMFT, BBT (all ρ > 0.75 pairwise) — motivates regularized regression.
5. **No sex or paresis-side effect** on funcUseRatio.
6. **One outlier** (C1P01, funcUseRatio > 1) should be investigated in a sensitivity analysis.

**What comes next:**

The Python notebook (`main.ipynb`) uses these findings to:

- Build a regularized machine learning pipeline (Linear, Ridge, Lasso) with LOOCV
- Identify the most important features (Lasso feature importance)
- Perform a **discrepancy analysis** to flag individual patients with learned non-use

Together, R (statistics) and Python (machine learning) cover the full pipeline from data exploration to clinically actionable insights.

