---
title: "Statistical Analysis — Predicting Arm Use After Stroke"
author: "Asad NADEEM"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    number_sections: true
---

# Introduction

This R notebook performs the **statistical validation** part of the project:

- Descriptive statistics on all variables
- Normality tests (Shapiro-Wilk)
- Correlation analysis (Spearman — non-parametric)
- Group comparisons (Mann-Whitney U tests)
- Visualizations (histograms, QQ plots, scatter plots, box plots)

The machine learning pipeline is in `main.ipynb` (Python). This notebook loads the merged dataset produced by Python (`data/merged_dataset.csv`).

---

# Setup

We load the libraries and the merged dataset.

```{r setup, include=FALSE}
# Knit options: control how chunks appear in the final HTML
knitr::opts_chunk$set(
  echo = TRUE,          # show R code in the output
  warning = FALSE,      # hide warnings
  message = FALSE,      # hide package loading messages
  fig.width = 10,       # default figure width
  fig.height = 6        # default figure height
)
```


```{r load-libraries-and-data}
# Required packages
library(readr)      # read CSV files
library(ggplot2)    # nice plots
library(dplyr)      # data manipulation
library(corrplot)   # correlation heatmaps
library(knitr)      # nice tables with kable()

# Load the merged dataset produced by the Python notebook
merged_df <- read_csv("data/merged_dataset.csv")

# Quick check
cat("Dataset:", nrow(merged_df), "patients,", ncol(merged_df), "columns\n")
cat("Columns:", paste(names(merged_df), collapse = ", "), "\n")
```
