---
title: "Statistical Analysis — Predicting Arm Use After Stroke"
author: "Asad NADEEM"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    number_sections: true
---

# Introduction

This R notebook performs the **statistical validation** part of the project:

- Descriptive statistics on all variables
- Normality tests (Shapiro-Wilk)
- Correlation analysis (Spearman — non-parametric)
- Group comparisons (Mann-Whitney U tests)
- Visualizations (histograms, QQ plots, scatter plots, box plots)

The machine learning pipeline is in `main.ipynb` (Python). This notebook loads the merged dataset produced by Python (`data/merged_dataset.csv`).

---

# Setup

We load the libraries and the merged dataset.

```{r setup, include=FALSE}
# Knit options: control how chunks appear in the final HTML
knitr::opts_chunk$set(
  echo = TRUE,          # show R code in the output
  warning = FALSE,      # hide warnings
  message = FALSE,      # hide package loading messages
  fig.width = 10,       # default figure width
  fig.height = 6        # default figure height
)
```


```{r load-libraries-and-data}
# Required packages
library(readr)      # read CSV files
library(ggplot2)    # nice plots
library(dplyr)      # data manipulation
library(corrplot)   # correlation heatmaps
library(knitr)      # nice tables with kable()

# Load the merged dataset produced by the Python notebook
merged_df <- read_csv("data/merged_dataset.csv")

# Quick check
cat("Dataset:", nrow(merged_df), "patients,", ncol(merged_df), "columns\n")
cat("Columns:", paste(names(merged_df), collapse = ", "), "\n")
```

# Descriptive Statistics

We compute the basic summary statistics (mean, standard deviation, median, min, max) for each continuous variable. These give us a first sense of the data: central tendency, spread, and range.

```{r descriptive-stats}
# List the continuous variables we want to summarize
continuous_vars <- c("Age", "PANU", "Barthel", "FMUE_TOTAL",
                     "WMFT_Total", "BBT_P", "funcUseRatio")

# Build the summary table row by row
summary_stats <- data.frame(
  Variable = continuous_vars,
  N        = sapply(continuous_vars, function(v) sum(!is.na(merged_df[[v]]))),
  Mean     = sapply(continuous_vars, function(v) round(mean(merged_df[[v]], na.rm = TRUE), 2)),
  SD       = sapply(continuous_vars, function(v) round(sd(merged_df[[v]], na.rm = TRUE), 2)),
  Median   = sapply(continuous_vars, function(v) round(median(merged_df[[v]], na.rm = TRUE), 2)),
  Min      = sapply(continuous_vars, function(v) min(merged_df[[v]], na.rm = TRUE)),
  Max      = sapply(continuous_vars, function(v) max(merged_df[[v]], na.rm = TRUE))
)

# Display as a formatted HTML table
kable(summary_stats, caption = "Descriptive statistics of continuous variables")
```

## Categorical Variables

For categorical variables (Sex, Paresis side), we just report counts.
```{r categorical-stats}
cat("Sex distribution:\n")
print(table(merged_df$Sex))

cat("\nParesis side distribution:\n")
print(table(merged_df$Paresis))
```
# Normality Tests (Shapiro-Wilk)

To decide between **Pearson** (parametric) and **Spearman** (non-parametric) correlations, we test each variable for normality.

**Shapiro-Wilk test rule:**

- p ≥ 0.05 → data is approximately normal → Pearson OK
- p < 0.05 → data is NOT normal → use Spearman

With a small sample (n = 30), Shapiro-Wilk is the recommended normality test.

```{r normality-tests}
normality_results <- data.frame(
  Variable = continuous_vars,
  W_statistic = NA,
  p_value = NA,
  Normal = NA
)

for (i in seq_along(continuous_vars)) {
  variable_name <- continuous_vars[i]
  values <- merged_df[[variable_name]][!is.na(merged_df[[variable_name]])]
  
  test_result <- shapiro.test(values)
  
  normality_results$W_statistic[i] <- round(test_result$statistic, 4)
  normality_results$p_value[i]     <- round(test_result$p.value, 4)
  normality_results$Normal[i]      <- ifelse(test_result$p.value >= 0.05, "Yes", "No")
}

kable(normality_results, caption = "Shapiro-Wilk normality test (alpha = 0.05)")
```

## QQ Plots — Visual Normality Check

A **QQ plot** shows the data quantiles against theoretical normal quantiles. If points align with the red diagonal line, the data is approximately normal. Deviations from the line indicate non-normality.

```{r qq-plots, fig.width=14, fig.height=10}
# Arrange 8 plots in a 2x4 grid
par(mfrow = c(2, 4))

for (variable_name in continuous_vars) {
  values <- merged_df[[variable_name]][!is.na(merged_df[[variable_name]])]
  
  qqnorm(values, main = variable_name,
         col = "steelblue", pch = 16)
  qqline(values, col = "red", lwd = 2)
}

# Reset to default single plot layout
par(mfrow = c(1, 1))
```

## Histograms — Distribution Shape

Histograms show how the data is spread out. We add a red dashed line at the mean for reference.

```{r histograms, fig.width=14, fig.height=8}
par(mfrow = c(2, 4))

for (variable_name in continuous_vars) {
  values <- merged_df[[variable_name]][!is.na(merged_df[[variable_name]])]
  
  hist(values,
       main = variable_name,
       xlab = variable_name,
       col = "steelblue",
       border = "black",
       breaks = 12)
  
  abline(v = mean(values), col = "red", lwd = 2, lty = 2)
}

par(mfrow = c(1, 1))
```

## Detailed Look at funcUseRatio

Since funcUseRatio is our target variable, we show it in detail with a histogram and a box plot side by side.

```{r target-detail, fig.width=12, fig.height=5}
par(mfrow = c(1, 2))

# Histogram with mean and median
hist(merged_df$funcUseRatio,
     breaks = 15,
     col = "steelblue",
     border = "black",
     main = "Distribution of funcUseRatio",
     xlab = "funcUseRatio")

abline(v = mean(merged_df$funcUseRatio), col = "red",    lwd = 2, lty = 2)
abline(v = median(merged_df$funcUseRatio), col = "orange", lwd = 2, lty = 2)

legend("topright",
       legend = c(paste("Mean =", round(mean(merged_df$funcUseRatio), 2)),
                  paste("Median =", round(median(merged_df$funcUseRatio), 2))),
       col = c("red", "orange"),
       lty = 2, lwd = 2, cex = 0.9)

# Box plot
boxplot(merged_df$funcUseRatio,
        main = "Box plot of funcUseRatio",
        ylab = "funcUseRatio",
        col = "steelblue")

par(mfrow = c(1, 1))

# Key observation
shapiro_p <- shapiro.test(merged_df$funcUseRatio)$p.value
cat("Shapiro-Wilk p-value:", round(shapiro_p, 4), "\n")
cat("Mean:", round(mean(merged_df$funcUseRatio), 3), "\n")
cat("Median:", round(median(merged_df$funcUseRatio), 3), "\n")
cat("Since mean > median and p < 0.05 -> right-skewed, not normal.\n")
cat("Decision: use Spearman correlations (non-parametric).\n")
```

# Spearman Correlation Analysis

Since funcUseRatio is not normally distributed, we use **Spearman** correlations. Spearman is rank-based, so it does not assume normality and is robust to outliers.

## Correlations with funcUseRatio

```{r spearman-correlations}
# Clinical variables to test against funcUseRatio
clinical_vars <- c("PANU", "FMUE_TOTAL", "WMFT_Total", "BBT_P", "Barthel", "Age")

correlation_results <- data.frame(
  Variable     = clinical_vars,
  rho          = NA,
  p_value      = NA,
  Significance = NA
)

for (i in seq_along(clinical_vars)) {
  variable_name <- clinical_vars[i]
  
  # Keep only rows where both variables are present
  valid_rows <- complete.cases(merged_df[[variable_name]], merged_df$funcUseRatio)
  
  test_result <- cor.test(merged_df[[variable_name]][valid_rows],
                          merged_df$funcUseRatio[valid_rows],
                          method = "spearman")
  
  correlation_results$rho[i]     <- round(test_result$estimate, 3)
  correlation_results$p_value[i] <- round(test_result$p.value, 4)
  correlation_results$Significance[i] <- ifelse(test_result$p.value < 0.001, "***",
                                         ifelse(test_result$p.value < 0.01,  "**",
                                         ifelse(test_result$p.value < 0.05,  "*", "ns")))
}

kable(correlation_results,
      caption = "Spearman correlations with funcUseRatio (*** p<0.001, ** p<0.01, * p<0.05)")
```

## Correlation Heatmap

A visual representation of all pairwise correlations between variables. Red = positive, blue = negative, intensity = magnitude.

```{r correlation-heatmap, fig.width=9, fig.height=8}
# Impute missing values globally for the heatmap display
# (imputation for modeling is done inside LOOCV folds in Python)
heatmap_df <- merged_df
heatmap_df$PANU[is.na(heatmap_df$PANU)] <- median(heatmap_df$PANU, na.rm = TRUE)
heatmap_df$Barthel[is.na(heatmap_df$Barthel)] <- median(heatmap_df$Barthel, na.rm = TRUE)

# Variables to include in the matrix
heatmap_vars <- c("PANU", "FMUE_TOTAL", "WMFT_Total", "BBT_P",
                  "Barthel", "Age", "funcUseRatio")

# Compute Spearman correlation matrix
correlation_matrix <- cor(heatmap_df[, heatmap_vars], method = "spearman")

# Draw the heatmap
corrplot(correlation_matrix,
         method = "color",
         type = "lower",              # show only lower triangle (upper is redundant)
         addCoef.col = "black",       # print the correlation value inside each cell
         tl.col = "black",            # axis text color
         tl.srt = 45,                 # rotate axis text 45 degrees
         number.cex = 0.8,            # size of numbers inside cells
         col = colorRampPalette(c("#2166AC", "white", "#B2182B"))(100),
         title = "Spearman Correlation Matrix",
         mar = c(0, 0, 2, 0))
```

# Scatter Plots

## Significant Predictors

We plot the three significant predictors against funcUseRatio. The outlier C1P01 is highlighted in red.

```{r scatter-significant, fig.width=14, fig.height=5}
par(mfrow = c(1, 3))

# Index of the outlier
outlier_idx <- which(heatmap_df$ID == "C1P01")

# WMFT
plot(heatmap_df$WMFT_Total, heatmap_df$funcUseRatio,
     pch = 16, col = "steelblue",
     xlab = "WMFT_Total", ylab = "funcUseRatio",
     main = "WMFT vs funcUseRatio")
points(heatmap_df$WMFT_Total[outlier_idx],
       heatmap_df$funcUseRatio[outlier_idx],
       col = "red", pch = 16, cex = 1.5)

# BBT
plot(heatmap_df$BBT_P, heatmap_df$funcUseRatio,
     pch = 16, col = "#27AE60",
     xlab = "BBT_P", ylab = "funcUseRatio",
     main = "BBT vs funcUseRatio")
points(heatmap_df$BBT_P[outlier_idx],
       heatmap_df$funcUseRatio[outlier_idx],
       col = "red", pch = 16, cex = 1.5)

# FMUE
plot(heatmap_df$FMUE_TOTAL, heatmap_df$funcUseRatio,
     pch = 16, col = "#E67E22",
     xlab = "FMUE_TOTAL", ylab = "funcUseRatio",
     main = "FMUE vs funcUseRatio")
points(heatmap_df$FMUE_TOTAL[outlier_idx],
       heatmap_df$funcUseRatio[outlier_idx],
       col = "red", pch = 16, cex = 1.5)

par(mfrow = c(1, 1))
```

## PANU — Not Significant

PANU shows no clear relationship with funcUseRatio — confirming the finding from Le Perf (2022).

```{r scatter-panu, fig.width=7, fig.height=5}
plot(heatmap_df$PANU, heatmap_df$funcUseRatio,
     pch = 16, col = "gray60",
     xlab = "PANU (%)", ylab = "funcUseRatio",
     main = "PANU vs funcUseRatio (not significant)")

points(heatmap_df$PANU[outlier_idx],
       heatmap_df$funcUseRatio[outlier_idx],
       col = "red", pch = 16, cex = 1.5)

legend("topright", legend = "C1P01 (outlier)",
       col = "red", pch = 16)
```

# Group Comparisons

We test whether funcUseRatio differs by Sex or Paresis side using the **Mann-Whitney U test** — the non-parametric equivalent of a t-test (doesn't assume normality).

```{r group-comparisons, fig.width=10, fig.height=5}
par(mfrow = c(1, 2))

# By Sex
boxplot(funcUseRatio ~ Sex,
        data = merged_df,
        col = c("coral", "steelblue"),
        main = "funcUseRatio by Sex",
        ylab = "funcUseRatio")

test_sex <- wilcox.test(funcUseRatio ~ Sex, data = merged_df)
cat("Mann-Whitney test by Sex:     W =", test_sex$statistic,
    ", p =", round(test_sex$p.value, 3), "\n")

# By Paresis side
boxplot(funcUseRatio ~ Paresis,
        data = merged_df,
        col = c("steelblue", "coral"),
        main = "funcUseRatio by Paresis Side",
        ylab = "funcUseRatio")

test_paresis <- wilcox.test(funcUseRatio ~ Paresis, data = merged_df)
cat("Mann-Whitney test by Paresis: W =", test_paresis$statistic,
    ", p =", round(test_paresis$p.value, 3), "\n")

par(mfrow = c(1, 1))
```

